---
title: "Asdia, predicting IBD from structure of the gut microbiome"
author: "Clovis Galiez"
date: "January 24, 2019"
output: html_document
---

```{r setup, include=FALSE}
set.seed(42)
knitr::opts_chunk$set(echo = TRUE)
```


We want to predict the IBD clinical status of patients given the abundance of microbial species living in their gut. curatedMetagenomicData package allows to retrieve a matrix of patients together with their gut's microbial composition. Each patient is assigned to a clinical status.

We focus here on the metagenomic study by Nielsen H.B. et al, published in 2014.


## Load/install packages
```{r libs, include=FALSE}
#install.packages('curatedMetagenomicData')
#source('http://bioconductor.org/biocLite.R')
#biocLite('phyloseq')
#library(phyloseq)
library(curatedMetagenomicData)
library(ggplot2)
```

## Load the data

```{r pressure, echo=FALSE}
###########################
# define study considered #
###########################
study = "NielsenHB_2014"
  # create output dir
output.dir = paste("./", study, sep = "")
dir.create(output.dir, showWarnings = F)

##########################
# extract taxonomic data #
##########################
db = "metaphlan_bugs_list.stool"
  #-----------#
  # load data #
  #-----------#
cat("\t-loading taxonomic data data\n")
data = curatedMetagenomicData( paste(study, db, sep = "."), dryrun = FALSE)
data = data[[1]]
  # get expression 
X = exprs(data)
X = t(X)
  # get covariates
meta = data@phenoData@data
status = meta$study_condition
status = factor(status)
names_of_species = colnames(X)
```

Take a look to your data and understand what's in there.
```{r}
load('ibdStart.RData')
# The most important variables for you are:
dim(X)
head(status)
head(names_of_species)

```

## Predicting IBD


### Identify important bacterial species

Find potential correlations between abundances of bacterial species and IBD.

```{r explainatory vars}
z_and_o = as.numeric((status == "IBD")) # 1 if the personn is ill 
o_and_z = abs(z_and_o - 1)
C <- c()
for (i in 1:1939)
  C[i] = cor(z_and_o, X[ ,i])^2


# Finding the more influent species
which.max(C)
names_of_species[7]

# Calculating the average quantity of this species among IBD and control
z_and_o%*%X[, 7]/sum(z_and_o)

o_and_z%*%X[, 7]/sum(o_and_z)

#On ordonne status selon la quantité de cette bactérie
status_o = z_and_o[order(X[, 7], decreasing = T)]
inversed_status = o_and_z[order(X[, 7], decreasing = T)]

# True positive rate
TPR = cumsum(status_o)/sum(z_and_o)
# False positive rate
FPR = cumsum(inversed_status)/sum(o_and_z)

#AUC ROC
plot(FPR, TPR)



```

What AUC ROC can you reach with the most correlated species?
the most correlated species are k__Bacteria and p__Actinobacteria. We will keep the first one.


Is it the presence or the abscence of this species that is a marker of IBD?

On average, the IBD have 5.998106 and the control have 2.13289 of this baterie. thus, the presence of this bacteria is a marker of IBD.

Are you satisfied? Check quickly with other people around and make a small litterature search.
Give a reason why this species is marker of the IBD, but may not be relevant for the predicting IBD.

Clean your data according to your findings in order to have relevant prediction from the abundance data.

What is the new best correlated species. What p-value it corresponds to?

Now you should realize that to get a biomarker of IBD, one needs not only the measure of 1 single species but the combination of several species.


## Learning a predictor




### Logistic regression


Use the package glmnet to use a logistic regression for classification.

Think beforehand of what potential trap you should you be aware of?
How can you avoid it?


```{r}
library(glmnet)

fit = glmnet(X, z_and_o, family = "binomial")
coef(fit, s=fit$lambda.1se)

cvfit = cv.glmnet(X, z_and_o, family = "binomial", type.measure = "class")
plot(cvfit)
summary(fit)


```

What species you find relevant for predicting IBD? Are they in accordance with the litterature?


### Bonus: Neural networks

#### Load/intsall keras package
```{r}
#install.packages("keras")
library(keras)
#install_keras()
```


#### Train a neural network using keras to predict IBD. 

Play around with the hyper parameters of your neural network that to predict the IBD status from the species abundances.

Ask yourself:
Is it overfitting?
Do you get any idea of which species is associated to IBD? 

